Learning how to act: making good decisions with machine learning

### 2017-07-09 thesis/thesis_map.tex
spelling, citation fixes

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (accross viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software benchmarking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\end{enumerate}

@@

The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}. This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

@@

The what-if type questions from the why. \cite{Gelman2010} \cite{}. Why do whites to better than blacks in school (in America). Suggests that reverse causal inference questions are more interesting and motivate most of social science.

@@

To explain or to predict \cite{Shmueli2010}
The two cultures \cite{Breiman2001}

@@

We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph. There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure. These rules are referred to as the Do-calculus \cite{Pearl2000}. We discuss them further in section \ref{sec:do_calculus}.

@@

The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables. 
By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph. This gives rise to the do-calculus \cite{Pearl2000}. The do calculus consists of three rules. They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.

@@

The question of non-parametric identifiability is solved. The do calculus is complete \cite{Shpitser2006b,Huang2006}. A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus. There is a polynomial time algorithm \cite{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:

@@

Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries. I have created a javascript implementation of the identifiability algorithm \cite{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.

@@

%Cyclic models \cite{Hyttinen2012}

@@

\subsection{Conditional independence based methods}
One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution. For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network. If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$. However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network. This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$. This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$. Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{pearl1988probabilistic} for $P$.

@@

\begin{theorem}
\cite{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
\end{theorem}

@@

For any DAG with latent (and selection) variables there is a unique MAG \cite{Richardson2002}. This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}. The logic behind the algorithm is very similar. Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations. The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \cite{Spirtes1995}. Assuming there are no selection variables, the PAG can contain four types of link:

@@

The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \cite{Zhang2008}. More recently \cite{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster. \cite{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.

@@

Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$. What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from? Recall the functional definition of causality (section \ref{sec:SEM}). There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a} or linear models with non-Gaussian noise \cite{Hoyer2012}.

@@

Randomisation does not ensure target and control group are exactly alike. The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables in a between the target and control groups in a finite data sample. However, the variance in the outcome within both the target and control groups also increases. The net result is increased variance (but not bias) in  the estimate of causal effects.

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest. If we stratify in such a way that the probability an instance recieves a given treatement is indepdenent of its covariates, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can still estiamte the average causal effects directly from the differences in outcome accross treatments. However, more complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}. The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between statified versus completely random experiments.

@@

Multi-armed bandits address the problem of desiging experiments that can adapt as samples are observed. Their introduction is generally attributed to \citet{Thompson1993}. In its classic formulation \cite{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution. \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated. It sufficient (but not nessessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action. The goal of the learner is to maximise the total reward they receive. This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate}
At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$. We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$. 
\end{definition}

The total reward a bandit algorithm/policy can expect to acheive depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cummulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning liturature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.}.
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
}

Both the rewards and the actions selected by the algorithm are random variables. The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}. The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is refered to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}. A stochastic bandit algorithm is learning if it obtains psuedo-regret that is sub-linear in $T$.

@@

The regret is invariant to adding a constant to the expected rewards for all actions. However, it still depends on key characteristics of the reward distributions for each action. Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e bernoulli bandits, gaussian bandits) or that the rewards are bounded in some range. Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).

@@

The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators. The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards. As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable. The expectation and variance guarentees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example. This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem. Instead, the focus is on computing bounds on the expected regret.

\begin{figure}[h]
\centering
\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples. This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim bernoulli(0.2)$. In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$. $\E{\hat{\mu}} = 0.439$. The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping. Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded. This has some interesting potential real world implications. Early stopping of clinical trials is controversial. A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
\label{fig:early_stopping}
\includegraphics[scale=1]{extreme_early_stopping}
\end{figure}

There are a few key principles that are used to guide the development of bandit algorithms. The simplest is to explicitly seperate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps. A common example in practice is unform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase. This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.

Another key approach is \emph{optimism in the face of uncertainty}. Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound. The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high. Thus selecting it yields either a good reward or useful information.

\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the expontial family. They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptopically and proved this is asymptopically efficient. However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm. \citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$. Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds. We now assume the rewards are bounded in $[0,1]$. The algorithm and regret bounds can be generalised to subgaussian reward distributions, see \citet{Bubeck2012}.

@@

Somewhat unintuitively, the regret increases as the value of the arms gets closer together. This is because it becomes harder for the algorithm to identify the optimal arm. As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them. The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$. The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}. By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.

@@

Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}. This yeilds regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.

@@

Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal. This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}. Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}. However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivelent to UCB.

\subsection{Pure-exploration problems}
Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}. In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$. The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.

@@

\vspace{0.5cm}
\begin{definition}[Adversarial k-armed bandit problem]
\label{def:adversarial-k-armed-bandit}
Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
\begin{enumerate}
\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate} 
\end{definition}

Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivous}, as opposed to \emph{non-oblivous} adversaries, which can generate rewards as a function of the history of the game. In the case of oblivous adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.

For oblivious adversarial bandits we can define regret analagously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also refered to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}. We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.

>>>>

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\end{enumerate}

@@

The notion of causality has been widely debated in science and philosophy \citep{} but is still widely viewed as poorly defined \citep{}. This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

@@

The what-if type questions from the why. \citep{Gelman2010} \citep{}. Why do whites to better than blacks in school (in America). Suggests that reverse causal inference questions are more interesting and motivate most of social science.

@@

To explain or to predict \citep{Shmueli2010}
The two cultures \citep{Breiman2001}

@@

We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph. There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure. These rules are referred to as the Do-calculus \citep{Pearl2000}. We discuss them further in section \ref{sec:do_calculus}.

@@

The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables. 
By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph. This gives rise to the do-calculus \citep{Pearl2000}. The do calculus consists of three rules. They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.

@@

The question of non-parametric identifiability is solved. The do calculus is complete \citep{Shpitser2006b,Huang2006}. A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus. There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:

@@

Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries. I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.

@@

%Cyclic models \citep{Hyttinen2012}

@@

\subsection{Conditional independence based methods}
One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution. For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network. If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$. However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network. This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$. This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$. Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.

@@

\begin{theorem}
\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
\end{theorem}

@@

For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}. This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites}. The logic behind the algorithm is very similar. Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations. The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}. Assuming there are no selection variables, the PAG can contain four types of link:

@@

The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}. More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster. \citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.

@@

Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$. What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from? Recall the functional definition of causality (section \ref{sec:SEM}). There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.

@@

Randomisation does not ensure target and control group are exactly alike. The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample. However, the variance in the outcome, within both the target and control groups, also increases. The net result is increased variance (but not bias) in  the estimate of causal effects.

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments. More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}. The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

@@

Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed. Their introduction is generally attributed to \citet{Thompson1993}. In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution. \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated. It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action. The goal of the learner is to maximise the total reward they receive. This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate}
At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$. We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$. 
\end{definition}

The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.}.
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
}

Both the rewards and the actions selected by the algorithm are random variables. The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}. The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}. A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.

@@

The regret is invariant to adding a constant to the expected rewards for all actions. However, it still depends on key characteristics of the reward distributions for each action. Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range. Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).

@@

The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators. The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards. As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable. The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example. This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem. Instead, the focus is on computing bounds on the expected regret.

\begin{figure}[h]
\centering
\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples. This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$. In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$. $\E{\hat{\mu}} = 0.439$. The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping. Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded. This has some interesting potential real world implications. Early stopping of clinical trials is controversial. A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
\label{fig:early_stopping}
\includegraphics[scale=1]{extreme_early_stopping}
\end{figure}

There are a few key principles that are used to guide the development of bandit algorithms. The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps. A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase. This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.

Another key approach is \emph{optimism in the face of uncertainty}. Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound. The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high. Thus selecting it yields either a good reward or useful information.

\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family. They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient. However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm. \citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$. Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds. We now assume the rewards are bounded in $[0,1]$. The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.

@@

Somewhat unintuitively, the regret increases as the value of the arms gets closer together. This is because it becomes harder for the algorithm to identify the optimal arm. As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them. The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$. The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}. By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.

@@

Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}. This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.

@@

Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal. This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}. Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}. However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.

\subsection{Pure-exploration problems}
Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}. In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$. The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.

@@

\vspace{0.5cm}
\begin{definition}[Adversarial k-armed bandit problem]
\label{def:adversarial-k-armed-bandit}
Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
\begin{enumerate}
\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate} 
\end{definition}

Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game. In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.

For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}. We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
